\addchap{Preface}
\begin{refsection}
This volume contains a selection of revised peer-reviewed papers from the 49th Annual Conference on African Linguistics, held at Michigan State University in 2018. About 170 papers on both descriptive and theoretical aspects of African linguistics were presented at the conference. Areas of interest included phonetics, phonology, morphology, syntax, semantics, sociolinguistics, historical linguistics, discourse analysis, language documentation, computational linguistics and more. Presenters were both graduate students and more senior scholars based in North America, Africa and beyond. Their research covered all major regions of Africa and most language families found in the continent. 

The volume editors would like to thank those who contributed to the success of the conference through their generous support and those who made the publication of this volume possible. These include the Association of Contemporary African Linguistics, the African Studies Center at Michigan State University (MSU) under the late Director James A. Pritchett and later under the stewardship of Jamie Monson, the MSU Department of Linguistics, Germanic, Slavic, Asian and African Languages (now Department of Linguistics, Languages, and Cultures) then chaired by Sonja Fritzche and later Jason Merrill, the College of Arts and Letters especially Dean Christopher P. Long,  Associate Dean Sonja Fritzche and Bill Hart-Davidson (Associate Dean for Research and Graduate Studies), and MSU’s Writing Center. Many thanks to Yen-Hwei Lin, the Chair of the Department of Linguistics, Languages, and Cultures for providing much-needed support. We would also like to express our gratitude to previous ACAL conference hosts who shared technical information with us, particularly UC Berkeley and Indiana University. Thanks also to our graduate and undergraduate student helpers, Foreign Language Teaching Assistants (FLTAs), tutors, administrative staff, and friends for their various roles in supporting the 49\textsuperscript{th} ACAL Conference. Tanner Schudlich, Nicole McKenzie, Nellie Hunter, Lisa Hinds, Michel Burton, Ok-Sook Park, Brahim Chakrani, Salamatu Abdulkarreem, Morgan Momberg, Magdalyne Oguti, Idris Abubakar, Felix Umeana, Emily Skupin, Yan Cong, Rachel Stacey, Adam Smolinski, and Yuakai Chen all deserve special mention.

\begin{sloppypar}
We also take this opportunity to thank many reviewers and proofreaders who took some time out of their busy schedules to help us select and improve the quality of the papers in this volume. Reviewers include Xiayimaierdan Abudushalamu, Gregory D. S. Anderson, Matthew Baerman, William G. Bennett, Lee Bickmore, Robert Botne, Kenyon Branan, Leston Buell, Mike Cahill, Yan Cong, Toni Cook, Katherine Demuth, Thabo Ditsele, Michael Diercks, Laura Downing, James Esegbey, Daniel Finer, Joash J. Gambarage, John Gluckman, John Goldsmith, Richard T. Griscom, Claire Halpert, Larry Hyman, Peter Jenks, Kyle Jerro, Jason Kandybowicz, Boniface Kawasha, Ettien Koffi, Ruth Kramer, Nancy C. Kula, Karsten Legère, Gastor C. Mapunda, Michael Marlo, Leonard Muaka, Laura McPherson, Steve Nicolle, David Odden, Kenneth S. Olson, Mary Paster, Doris Payne, Asia Pietraszko, Douglas Pulleyblank, Ronald Schaefer, Sharon Rose, Josephat Rugemarila, Ken Safir, Hannah Sande, Sylvester R. Simango, Jenneke van der Wal, Aggrey Wasike, Jochen Zeller and Patricia Schneider-Zioga.
\end{sloppypar}

The volume editors also express their gratitude to the Contemporary African Linguistics series editors Michael Marlo, Laura Downing, and Akinbiyi Akinlabi for providing support and direction. At Language Science Press, we greatly appreciate all the assistance we received and the quality of work from Sebastian Nordhoff, Felix Kopecky and other members of their team. In short the volume editors are grateful to those who provided the necessary support, presenters, invited speakers, conference audiences, contributors to this volume, publishers, and all those who worked behind the scenes.

% % {\sloppy\printbibliography[heading=subbibliography]}
\end{refsection}
